---
title: "Extracting tiles and annotations from an orthomosaic to train a CNN"
author: "Etienne Laliberté"
date: "17/11/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      root.dir = '/processing/cnn')
```

## Background

Convolutional neural networks (CNN) can be used to map plant species from high-resolution RGB drone imagery. A typical workflow is to acquire imagery and create RGB orthomosaics from it, and then use this imagery as a basemap to create annotations in vector format (e.g. polygons outlining individual plants, identified to species).

This tutorial that the type of CNN used is one of **semantic segmentation**. It could be adapted, however, for instance segmentation since in this dataset, individual trees have been identified to species as individual polygons from the imagery. 

To train such a CNN, one would usually start with:

* a high-resolution RGB orthomosaic (e.g. GeoTiff)
* a polygon layer of species annotations (e.g. a shapefile or other vector format)

The goal is to create:

* tiles of fixed size (e.g. 128 x 128 pixels) with the RGB information (e.g. three-band raster, as GeoTiff)
* masks of annotations for the same area (e.g. the same region of 128 x 128 pixels) with the species information (e.g. single-band raster, as GeoTiff)

## Import the orthomosaic
For this example we will use a cropped region of a high-resolution orthomosaic for a larger region of the Station de biologie des Laurentides (SBL). The imagery was acquired on October 7 2021, during peak fall leaf colors.

> Note: the GeoTiff has four bands (band 4 is transparency, bands 1:3 are RGB). We only want the RGB bands.

```{r ortho-view, message=FALSE}
library(stars) # library to deal with rasters
bands_sel <- 1:3 # bands to select (RGB only)
img <- read_stars("ortho_crop.tif", # read the GeoTiff
                    proxy = TRUE) %>%  # import to not load the imagery in memory
  .[,,, bands_sel] # select only RGB
plot(img, rgb = 1:3, main = "RGB") # plot RGB orthomosaic (low res version)
```

## Import the annotations
The tree annotations are stored in a GeoPackage file. We load the data and plot it.

```{r annot-import}
species <- read_sf("trees_crop.gpkg") # read the GeoPackage containing the tree polygons ID'ed to species
plot(species["Label"], main = "Classes") # plot the tree polygons
```

## Get species (and other classes) names
We now need to store a table listing all classes (mostly species), and create a numeric column listing a number for each class. This information needs to be kept since this will be our legend enabling us to translate numbers to species later on.
```{r sp-names, message=FALSE}
# Get species names
library(tidyverse)
sp_names <- species %>% 
  st_drop_geometry() %>% # transform to tibble
  select(Label) %>% # keep only the Label colum
  distinct() %>% # find the unique values
  arrange(Label) %>% # sort in alphabetical order
  mutate(sp_id = as.numeric(as.factor(Label))) # create new column with number IDs for each class (mostly species)
sp_names # show the result
```

We also edit the `species` layer to only keep one column with the `sp_id`. The reason is that CNN want numbers.

```{r sp-id}
species <- species %>%
 transmute(sp_id = as.numeric(as.factor(Label))) # we keep only one column with species numbers IDs
species # we inspect it
```

## Create the grid for the tiles
We need to create a grid to create the tiles that will be used to extract the RGB data and masks.

### Find the spatial resolution of imagery
 The first step is to know the resolution of the imagery, because we need to know the distance that corresponds to the number of pixels we want to extract.
```{r get-res}
img_xres <- abs(st_dimensions(img)[[1]]$delta) # x resolution
img_yres <- abs(st_dimensions(img)[[2]]$delta) # y resolution

# Ensure that pixels are square
round(img_xres, 3) == round(img_yres, 3) # should be TRUE
```

### Define the tile size
This is an important parameter. If the CNN expect tiles 128 x 128 pixels, then this parameter should be set to 128.
```{r n-pixels}
n_pixels <- 128 # tile size (width/height)
```

We must now calculate, in meters, the size of a 128 x 128 tile (or cell) in this imagery.
```{r tile-size}
cell_size <- n_pixels * img_xres # 128 pixels x the size of one pixel
cell_size # 2.944 m x 2.944 m in this case
```

> Note: 3 m x 3 might be a bit small to train this CNN, given the size of the large trees. But we keep going anyway as this is simply for illustration purposes.

Then, we make the grid. And we plot it over the RGB imagery to check it.
```{r make-grid}
aoi_grid <- st_make_grid(img, # this uses the extent of the imagery
                         cellsize = cell_size) # we make cells the size of our 128x128 tiles
plot(img, rgb = 1:3, main = "RGB", reset = FALSE) # we plot the imagery
plot(aoi_grid, add = TRUE) # we overlay the grid
```

We will avoid cells which are not entirely within the bounding box of the imagery. This can be done with the following code.
```{r within}
tiles_within <- st_within(aoi_grid, st_as_sfc(st_bbox(img))) # finds cells within the extent of the imagery
sel_logical = lengths(tiles_within) > 0 # returns of logical vector for those cells we keep
aoi_grid_sel <- aoi_grid[sel_logical] # we filter only those cells
```

We see the some cells at the northern and eastern margins are now excluded.
```{r cell-excl, echo=FALSE}
plot(img, rgb = 1:3, main = "RGB", reset = FALSE)
plot(aoi_grid_sel, add = TRUE)
```

## Extract the tiles and masks
We are now ready to extract the tiles and masks. To speed things up, we will run this in parallel on multiple cores.

### Preparing for parallel computing
We need to prepare everything for parallel computing. This implies telling **R** how many cores to use. We use all but one, to ensure that we still have one core that remains free for other stuff, if needed. But you could decide to use all available cores to speed things up further.
```{r prep-paral, message=FALSE, eval=FALSE}
library(foreach)
library(doParallel)
n_cores <- detectCores() - 1 # we use all cores but one
cl <- makeCluster(n_cores) # we start the parallel cluster
registerDoParallel(cl) # we register it
clusterExport(cl, "bands_sel") # we need to export that object to all cores
```

Before starting the loop, one parameter to define is the bounding box shrink factor for grid cell size to extract tiles. We set it to 0.0001. See `?st_crop` for details. 
```{r crop-eps}
crop_eps <- 0.0001
```

### First loop: list of cropped tiles
Then we run the first parallel loop. That first loop generates for every tile (or cell) in our grid `aoi_grid_sel`, a 128 x 128 pixel RGB tile from the imagery. The list will be as long as there are cells in our grid. To be clear, this first step just creates the metadata and stores it as a lit of `stars_proxy` objects.

```{r grid-loop, eval=FALSE}
# Loop to extract tile locations (as stars_proxy objects)
tiles_stars_proxy <- foreach(i = 1:length(aoi_grid_sel),
                             .packages = c('tidyverse', 'stars')) %dopar% {
                               
    # Extract tile i from img
    tile_i <- img %>% 
      st_crop(aoi_grid_sel[i],
              epsilon = crop_eps)
    
} # end of foreach loop for tile locations
```

### Second loop: save the RGB tiles

Then, we use this list and the second loop actually creates the very many 128 x 128 RGB tiles extracted from the orthomosaic. They are stored as GeoTiff files.

> Make sure that folders tiles `tiles` and `masks` are created first in your working directory. These are used to store the outputs.

```{r tile-exc, eval=FALSE}
tiles <- foreach(i = 1:length(tiles_stars_proxy),
                 .packages = c('tidyverse', 'stars')) %dopar% {
  
  # Extract tile i
  tile_i <- tiles_stars_proxy[[i]] %>% 
    # write GeoTiff
    write_stars(paste0("tiles/tile_", str_pad(i, 4, pad = "0"), ".tif"),
    type = "Byte",
            options = c("COMPRESS=LZW"))
}
```

### Third loop: extract the masks

Finally, we save the masks. We first need to set a background (class corresponding to no annotations). We give it a value of 0 since our species (and other classes) started at 1.
```{r background-class}
bckgnd_class <- 0
```

This loop then crops the `species` layer in small tiles and converts the classes to a single-band raster. All of these 128 x 128 pixel masks are then stored in a folder.
```{r masks, eval=FALSE}
masks <- foreach(i = 1:length(aoi_grid_sel),
                 .packages = c('tidyverse', 'stars')) %dopar% {
                   
    # Extract species cell i from polygon layer             
    suppressWarnings(
        species_i <- species %>% 
           st_intersection(aoi_grid_sel[i]) )

    # Create one-band template for rasterization
    template <- tiles_stars_proxy[[i]] %>%
      st_as_stars() %>% 
      slice(band, 1)   
    
    # Set values to background class
    template[[1]][] <- bckgnd_class
    
    # Rasterize the species vector tile
    mask_i <- st_rasterize(species_i[, 'sp_id'], 
                            template) %>%
    # Export as GeoTiff
    write_stars(paste0("masks/mask_", str_pad(i, 4, pad = "0"), ".tif"),
    type = "Byte",
            options = c("COMPRESS=LZW"))
} # end foreach loop masks
```

Finally, we stop the cluster.
```{r stop-cluster, eval=FALSE}
stopCluster(cl)
```

We now have very many tiles and masks ready to be used to train a semantic segmentation CNN!

## Inspect the tiles and masks
The individual tiles and masks can be opened in QGIS or other GIS software to check that everything worked well. But we can just check a random tile and mask (we select the 22nd one).
```{r check-tile}
num <- 22
tile_num <- read_stars(paste0("tiles/tile_", str_pad(num, 4, pad = "0"), ".tif"))
plot(tile_num, rgb = 1:3)
```

Finally we can look at the associated mask for that tile.
```{r check-mask}
mask_num <- read_stars(paste0("masks/mask_", str_pad(num, 4, pad = "0"), ".tif"))
plot(mask_num, useRaster = FALSE)
```



